Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2023-01-24 text/Manuscript.Rmd
edit all figures

The ability to perceive changes in the environment and adapt behaviour accordingly is essential for the survival of all living organisms. Avoidance from UV radiation, which damages DNA essential for survival, is proposed to be one of the behaviours that existed before the abundant development of Earth's oxygen to block UV radiation in the atmosphere (Pittendrigh). In the ocean, almost all zooplankton have a pronounced rhythmic behaviour called Diel vertical migration (DVM), which consists of upwards towards the surface at dusk and descend back to deeper water before dawn. Avoidance behaviour to escape harmful UV radiation during the day is closely linked to the control of vertical migration in the ocean. However, little is known about the molecular neural circuit mechanisms in UV avoidance behaviour in many animals, including zooplankton. In neural circuits in general light response, the activation of various receptors by excitatory and inhibitory neurotransmitters such as glutamate, GABA and dopamine is known in various species. In addition to this, it has long been known that in nitric oxide (NO), as a neuromodulatory molecule, it is also important as a modulator of visual activity. NO is a small free-radical molecule with diffusible and changeable characteristics, and thus works as one of the neural signalling molecules. The function of NO in the neural circuitry of vision has been well studied in vertebrates. In the mammalian retina, NO is an essential signalling molecule in all retinal cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one of the retinal cells, activates soluble guanylate cyclase (sGC) expressed in cone bipolar cells, which produces cGMP and regulates downstream signalling. We therefore hypothesise that NO-mediated neural signalling may be involved in the photoreceptor circuits that control UV avoidance behaviour. To test this hypothesis we used the annelid *Platynereis dumerilii*, which conserved vertebrate-type ciliary photoreceptors (cPRCs) [@Arendt2004]. The cPRC expresses c-opsin1, which absorbs UV radiation, is involved in melatonin production and regulates circadian swimming behaviour [@Tosches2014; @tsukamoto2017]. Furthermore, we previously reported abnormal UV avoidance behaviour in c-opsin1 knockout larvae [@Veraszto2018]. The cPRCs express the UV-absorbing c-opsin1 and mediate UV avoidance in the larvae [@Veraszto2018]. In adult worms, UV light and c-opsin1 modulate locomotion and maturation [@Veedin2021]. These suggest that Platynereis larvae are a highly suitable model for elucidating the neural circuit mechanisms of NO-induced UV avoidance behaviour, and the accessibility of genome editing, behavioural assays, calcium imaging and whole-body connectomics could be very useful for study at the molecular level[@ozpolat2021].

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences (Figure 1---figure supplement 1). To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1A-D). The same four cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene (Figure 1E). This reporter also revealed the axonal projections of the NOS-expressing neurons. The position and morphology of the four *NOS*-expressing cells allowed us to identify the same cells in the three-day-old whole-body volume EM data [@williams2017; @veraszto2020] as four interneurons (INNOS)(Figure 1F and G). The INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Figure 1C, D). INRGW neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the prototroch ciliary band and the cholinergic MC ciliomotor neurons (Figure 1C, D). We also detected *NOS* expression by HCR in the region of the visual eyes and the pigmented eyespots.

![**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** SEM image of a 3-day-old Platynereis larva. **(B, C)** Five types of neurons (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed in the whole-body transmission electron microscopy (ssTEM) dataset of a 3-day-old larva. Ciliated cells labelled with (B) are shown in grey. The INNOS on the ventral side labelled with (C) are shown in purple. Axons and dendrites appear as lines and cell body positions are represented by spheres. **(D)** Expression analysis of the NOS gene (magenta) using in situ HCR. Anterior view of the larva at 2-day-old. Co-stained image using acetylated α-tubulin antibody (acTub: green). **(E)** Immunostaining with anti-HA antibody against HA tagged palmitoylated-tdTomato expressed under the upstream of the start site of NOS (NOSp::palmi-3xHA: magenta). Co-staining with acetylated α-tubulin antibody as marker for cilia of cPRC and axonal scaffold (acTub: green). Views of larvae at 3-day-old from the anterior side. Insert: One of the reconstracted INNOS is shown from the anterior side. **(F)** Wiring diagram of the cPRC circuit. Hexagons and arrows represent cell groups and their synaptic connections, respectively. Numbers inside hexagons indicate the number of cells grouped in each node. Arrow line thickness (synapse weight) is equal to the number of synapses.](../figures/Fig1.png){width="100%"}

@@

The specific expression of NOS in interneurons directly postsynaptic to the cPRCs suggests that NO signalling is involved in the UV response. To test this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to an early stop codon and thus likely representing null alleles (Figure 2A). We could establish a homozygous line for both mutations indicating that NOS is not an essential gene. To quantify UV avoidance, we recorded the trajectories of freely swimming larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm violet light. We also tested phototaxis, by exposing larvae to 480 nm directional and collimated light from the top of the column. As previously shown, wild-type larvae swim downward within 30 sec following non-directional UV/violet light stimulation [@Veraszto2018]. In contrast, homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 2B and C). This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour [@Veraszto2018]. *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function in the visual eyes mediating phototaxis [@Randel2014a], where *NOS* is also expressed (Figure 1). To test if the observed defects are due to a role of NOS in neuronal signalling or circuit developmental, we also tested larvae exposed to a NOS inhibitor, L-NAME. Larvae incubated for xy min in 0.1 mM or 1 mM L-NAME showed defective UV avoidance, but normal phototaxis. Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system.

To analyse the behaviour of larvae before and after UV stimulation in more detail, the temporal changes in some parameters were compared using tracking of individual larvae of the WT and mutant (Figure 2D). The results showed that in the WT there was a rapid and strong increase in downward swimming approximately 15 s after UV stimulation, whereas this did not occur in the mutant (Figure 2E). Interestingly, there was already a difference in swimming speed between the WT and the mutant before UV stimulation (in the dark) (Figure 2F). After UV stimulation, the swimming speed of the WT increased and reached a maximum after about 15 s. In contrast, the swimming speed of the mutant did not increase after about 15 s. In contrast, the mutants showed a slight increase in speed after UV stimulation, but swimming speed was not maintained and returned to the speed before UV stimulation. To investigate the role of NO on UV avoidance behaviour, receptors for NO were explored.

![**Figure 2. NO produced by UV stimulation to cPRCs.** **(A)** A NO-positive cell visualised by DAF-FM (green) is shown (dashline). The white line indicates the outer frame of the larva. All subsequent images are of 2-day-old larvae. **(B)** Intensity changes before (upper) and during 405 nm light stimulation (lower) in a cell visualised by DAF-FM. **(C)** The intensity of DAF-FM changes over time when cPRC (green) and other cells (ctr: gray) are stimulated with UV/violet (405 nm) light. Purple squares indicate the timing of 405 nm light stimulation, the transparent lines show the optical responses to individual intensities, each intensities are normalized (ΔF/F0) and the dark trace lines and the shaded bands show the mean value and the standard error range of the mean.](../figures/Fig2.png){width="100%"}

To test if NO is produced during UV/violet responses, we injected zygotes with the fluorescent NO-reporter DAF-FM and imaged two-day-old larvae during the local stimulation of the cPRC cilia by 405 nm light (Figure 4---figure supplement 1). Following light stimulation, we detected an increase in DAF-FM fluorescence in the region of the INNOS cells (Figure 4A). We next analysed the dynamics of the cPRC circuit following light stimulation. As we have shown previously [@Veraszto2018], a 20-sec illumination of the cPRC ciliary region lead to a decrease in cPRC calcium levels, likely due to the hyperpolarisation of the cPRCs. After 20-sec, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus - a response that may involve depolarisation. This activation phase occurs after the end of the stimulation and may be due to a slower neuroendocrine feedback [@Veraszto2018]. To determine whether NO mediates such a feedback, we measured calcium signals in *NOS*-mutant larvae. While we detected the inhibition phase in cPRC in *NOS*-mutants, this was not followed by activation and calcium levels remained low (Figure 4B and C). These results revealed an essential requirement for NOS signalling in the second phase of the cPRC response.

![**Figure 3. Larvae with inhibited NOS have a delayed response to UV-avoidance.** **(A)** Top: The domain and Exon/Intron structure of NOS. Bottom: Close-up region showing the genomic locus of NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target sites. Gray shows PAM sequences, red shows stop codons. **(B)** Overlaid trajectories for WT (left, n=20) and NOS mutant (right, n=31) 3-day-old larvae. Grey shows the trajectories before UV stimulation and purple shows the trajectories after UV stimulation. **(C)** The temporal changes in the vertical position of the WT and mutant larvae before and after UV stimulation are shown. The starting points of each larval trajectory are set to 0. After UV stimulation is indicated by purple squares. **(D)** Vertical swimming in wild-type (WT) and mutant (NOSΔ11 and NOSΔ23) larvae at 3-day-old stimulated with UV (395 nm) light from side, blue (488 nm) light from top and UV (395 nm) light from top. The data are shown in 30 s bins. **(E)** Vertical swimming in larvae treated with NOS inhibitors, L-NAME at 3-day-old stimulated with UV (395 nm) light from side, blue (488 nm) light from top and UV (395 nm) light from top. The data are shown in 30 s bins.](../figures/Fig3.png){width="100%"}

**Two unconventional guanylyl cyclases (NIT-GC) are expressed in the cPRCs**

NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PF00211). NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production. Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit ([@Veraszto2017] and data not shown). Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020]. To search for NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs. To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan species and [\*\*how many?] choanoflagellate. We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup. In agreement with Moroz et al. [@Moroz2020],we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans. The 15 *Platynereis* sequences were parts of several major and ancient clades of the tree (Figure 4---figure supplement 1). To characterise the expression of NIT-GCs, we using previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017]. Among the 15 NIT-GCs, four showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 3---figure supplement 1). In the single-cell data, we could identify the cPRCs by the specific expression of c-opsin1 and a pedal-peptide2 neuropeptide precursor (MLD proneuropeptide) [@Arendt2004; @williams2017], which have previously been described as cPRC markers (Figure 3A and B, Figure 3---figure supplement 2). The INNOS cells were identified by NOS expression and their spatial mapping. We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequences, NIT-GC1 and NIT-GC2. To confirm the single-cell data, we first carried out HCR with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes showed very clear and specific expression in the four cPRCs, as confirmed by co-labeling with an acetylated tubulin antibody and with an HCR probe against *MLD proneuropeptide*. To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins. In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus. Co-immunostaining with the rabbit NIT-GC1 and the rat NOS antibodies revealed that both proteins were localised in close proximity in the neurosecretory plexus (Figure 3---figure supplement 3). In contrast, NIT-GC2 strongly and specifically labelled the large sensory ciliary region of the cPRCs (Figure 3C and D). These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

To characterise the role of NIT-GCs within the cPRC during UV stimulation in more detail, the calcium sensor GCaMP6s were ubiquitously expressed in larvae in which NIT-GCs were knocked down using morpholino to examine changes in calcium levels in the cPRC. Local stimulation of the ciliated part of the cPRC with 405 nm light causes a transient decrease in calcium levels in the cPRC (hyperpolarisation), followed by an increase in calcium levels (depolarisation) when the light is switched off [@Veraszto2018]. In NIT-GC1 morphant larvae, 405 nm light stimulation caused a decrease in calcium levels in the cPRC, but the increase in calcium levels that occurred after the light stimulation was stopped was strongly suppressed (Figure 3E). This result suggests that NIT-GC1 mediates depolarising responses after UV stimulation of cPRC and may be involved in extracellular signalling. Interestingly, in NIT-GC2 morphant larvae, an increase in calcium levels occurred immediately after light stimulation (Figure 3F). The different responses of cPRC upon knockdown of NIT-GC1 and NIT-GC2 suggest that there may be different intracellular signalling mechanisms associated with their respective localisations. We focused our analysis on NIT-GC1 because it was localised to cPRC neuroaxons within the neuroplexas region and is involved in extracellular signalling after UV stimulation. It is generally known that depolarisation in the PRC involves elevated cGMPs in the photoreceptors. We therefore used an in vitro system in which cultured cells expressed NIT-GC1 to assess the cGMP activity of NIT-GC1.

@@

NIT-GC1 has a highly conserved cyclase domain and activation of NIT-GC1 is expected to produce cGMP. Recently, a synthetic protein called Green cGull was developed as a cGMP indicator that can be used in cultured cell systems [@Matsuda2017]. We co-expressed Green cGull and NIT-GC1 using the 2A self-cleaving peptide in cultured cells COS-7 (monkey kidney) and tested the intracellular dynamics of cGMP in the activation of NIT-GC1 (Figure 3---figure supplement 4). When the NO donor SNAP was applied to the NIT-GC1 and Green cGull co-expression cells, the intracellular cGMP levels was significantly increased compared to when NIT-GC1 was not expressed (Figure 3G and H). The increase in fluorescence intensity also did not occur when DMSO was applied (Figure 3I). To confirm whether this increase in intracellular cGMP levels is NIT domain-dependent response of NIT-GC1, mutant form of NIT-GC1 without the NIT domain was expressed and SNAP was applied (Figure 3---figure supplement 4A). As a result, the increase in fluorescence intensity was significantly reduced (Figure 3J). These results indicate that NIT-GC1 produces cGMP via the NIT domain. In bacteria, it has been proposed that the NIT domain regulates cellular functions in response to changes in extracellular and intracellular nitrate and nitrite concentrations. Our results suggest that NIT-GC1 may be sensitive to NO. This could be a possible feedback pathway in which NIT-GC1 expressed in cPRC is activated by NO produced by IN-NOS. In order to further explore the role of NO in the cPRC circuit, we therefore compared calcium imaging of postsynaptic cells (IN-NOS, IN-RGW, Ser-h1, MC) in the cPRC circuit during UV stimulation of cPRC between WT and NOS mutant larvae.

>>>>

The ability to perceive changes in the environment and adapt behaviour accordingly is essential for the survival of all living organisms. Avoidance from UV radiation, which damages DNA essential for survival, is proposed to be one of the behaviours that existed before the abundant development of Earth's oxygen to block UV radiation in the atmosphere (Pittendrigh). In the ocean, almost all zooplankton have a pronounced rhythmic behaviour called Diel vertical migration (DVM), which consists of upwards towards the surface at dusk and descend back to deeper water before dawn. Avoidance behaviour to escape harmful UV radiation during the day is closely linked to the control of vertical migration in the ocean. However, little is known about the molecular neural circuit mechanisms in UV avoidance behaviour in many animals, including zooplankton. In neural circuits in general light response, the activation of various receptors by excitatory and inhibitory neurotransmitters such as glutamate, GABA and dopamine is known in various species. In addition to this, it has long been known that in nitric oxide (NO), as a neuromodulatory molecule, it is also important as a modulator of visual activity. NO is a small free-radical molecule with diffusible and changeable characteristics, and thus works as one of the neural signalling molecules. The function of NO in the neural circuitry of vision has been well studied in vertebrates. In the mammalian retina, NO is an essential signalling molecule in all retinal cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one of the retinal cells, activates *soluble guanylate cyclase (sGC)* expressed in cone bipolar cells, which produces cGMP and regulates downstream signalling. We therefore hypothesise that NO-mediated neural signalling may be involved in the photoreceptor circuits that control UV avoidance behaviour. To test this hypothesis we used the annelid *Platynereis dumerilii* **(Figure 1A)**, which conserved vertebrate-type ciliary photoreceptors (cPRCs) [@Arendt2004]. The cPRC expresses *c-opsin1*, which absorbs UV radiation, is involved in melatonin production and regulates circadian swimming behaviour [@Tosches2014; @tsukamoto2017]. Furthermore, we previously reported abnormal UV avoidance behaviour in *c-opsin1* knockout larvae [@Veraszto2018]. The cPRCs express the UV-absorbing c-opsin1 and mediate UV avoidance in the larvae [@Veraszto2018]. In adult worms, UV light and c-opsin1 modulate locomotion and maturation [@Veedin2021]. These suggest that Platynereis larvae are a highly suitable model for elucidating the neural circuit mechanisms of NO-induced UV avoidance behaviour, and the accessibility of genome editing, behavioural assays, calcium imaging and whole-body connectomics could be very useful for study at the molecular level[@ozpolat2021].

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences **(Figure 1---figure supplement 1)**. To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region **(Figure 1D and Figure 1---figure supplement 2)**. The same four cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene **(Figure 1E)**. This reporter also revealed the axonal projections of the NOS-expressing neurons. The position and morphology of the four *NOS*-expressing cells allowed us to identify the same cells in the three-day-old whole-body volume EM data [@williams2017; @veraszto2020] as four interneurons (INNOS) **(Figure 1B,C)**. The INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets **(Figure 1C,F)**. INRGW neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the prototroch ciliary band and the cholinergic MC ciliomotor neurons **(Figure 1C,F)**. We also detected *NOS* expression by HCR in the region of the visual eyes (adult eyes) and the pigmented eyespots **(Figure 1---figure supplement 1)**.

![**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** SEM image of a three-day-old Platynereis larva. **(B, C)** Five types of neurons (cPRC, INNOS, INRGW, Ser-h1 and MC) in the cPRC circuit reconstructed in the whole-body transmission electron microscopy (ssTEM) dataset of a three-day-old larva. Ciliated cells labelled with (B) are shown in grey. The INNOS on the ventral side labelled with (C) are shown in purple. Axons and dendrites appear as lines and cell body positions are represented by spheres. **(D)** Expression analysis of the NOS gene (magenta) using in situ HCR. Anterior view of the larva at two-day-old. Co-stained image using acetylated α-tubulin antibody (acTub: green). **(E)** Immunostaining with anti-HA antibody against HA tagged palmitoylated-tdTomato expressed under the upstream of the start site of NOS (NOSp::palmi-3xHA: magenta). Co-staining with acetylated α-tubulin antibody as marker for cilia of cPRC and axonal scaffold (acTub: green). Views of larvae at three-day-old from the anterior side. Insert: One of the reconstracted INNOS is shown from the anterior side. **(F)** Wiring diagram of the cPRC circuit. Hexagons and arrows represent cell groups and their synaptic connections, respectively. Numbers inside hexagons indicate the number of cells grouped in each node. Arrow line thickness (synapse weight) is equal to the number of synapses.](../figures/Fig1.png){width="100%"}

The specific expression of *NOS* in interneurons directly postsynaptic to the cPRCs suggests that NO signalling is involved in the UV response. To test if NO is produced during UV/violet responses, we injected zygotes with the fluorescent NO-reporter DAF-FM and imaged two-day-old larvae during the local stimulation of the cPRC cilia by 405 nm light. Following light stimulation, we detected an increase in DAF-FM fluorescence in the region of the INNOS cells **(Figure 2)**.

![**Figure 2. NO produced by UV/violet stimulation to cPRCs.** **(A)** A NO-positive cell visualised by DAF-FM (green) is shown (dashline). The white line indicates the outer frame of the larva. All subsequent images are of two-day-old larvae. **(B)** Intensity changes before (upper) and during 405 nm light stimulation (lower) in a cell visualised by DAF-FM. **(C)** The intensity of DAF-FM changes over time when cPRC (green) and other cells (ctr: gray) are stimulated with UV/violet (405 nm) light. Purple squares indicate the timing of 405 nm light stimulation, the transparent lines show the optical responses to individual intensities, each intensities are normalized (ΔF/F0) and the dark trace lines and the shaded bands show the mean value and the standard error range of the mean.](../figures/Fig2.png){width="100%"}

@@

To test this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to an early stop codon and thus likely representing null alleles **(Figure 3A)**. We could establish a homozygous line for both mutations indicating that NOS is not an essential gene. To quantify UV avoidance, we recorded the trajectories of freely swimming larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm UV/violet light. To analyse the behaviour of larvae before and after UV stimulation, the temporal changes in some parameters were compared using tracking of individual larvae of the WT and mutant **(Figure 3B,C and Figure 3---figure supplement 1A-D)**. The results showed that in the WT there was a rapid and strong increase in downward swimming approximately 15 s after UV stimulation, whereas this did not occur in the mutant **(Figure 3C and Figure 3---figure supplement 1B)**. Interestingly, there was already a difference in swimming speed between the WT and the mutant before UV stimulation (in the dark) **(Figure 3---figure supplement 1C,D)**. After UV stimulation, the swimming speed of the WT increased and reached a maximum after about 20 s. In contrast, the mutants showed a slight increase in speed after UV stimulation, but swimming speed was not maintained and returned to the speed before UV stimulation. We also tested phototaxis, by exposing larvae to 480 nm directional and collimated light from the top of the column. As previously shown, wild-type larvae swim downward within 30 sec following non-directional UV/violet light stimulation [@Veraszto2018]. In contrast, homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response **(Figure 3D and Figure 3---figure supplement 1E)**. This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour [@Veraszto2018]. *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function in the visual eyes mediating phototaxis [@Randel2014a], where *NOS* is also expressed **(Figure 1---figure supplement 1A,B)**. To test if the observed defects are due to a role of NOS in neuronal signalling or circuit developmental, we also tested larvae exposed to a NOS inhibitor, L-NAME. Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed defective UV avoidance, but normal phototaxis **(Figure 3E)**. Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system. To investigate the role of NO on UV avoidance behaviour, receptors for NO were explored.

![**Figure 3. Larvae with inhibited NOS have a delayed response to UV-avoidance.** **(A)** Top: The domain and Exon/Intron structure of NOS. Bottom: Close-up region showing the genomic locus of NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target sites. Gray shows PAM sequences, red shows stop codons. **(B)** Overlaid trajectories for WT (n=32) and NOS mutant (NOSΔ11/Δ11, n=26 and NOSΔ23/Δ23, n=47) at three-day-old larvae. 0 sec as the starting point. After 10 sec, UV (395 nm) stimulation from the side. **(C)** The temporal changes in the vertical position of the WT and mutant larvae before (dark) and after UV stimulation (side 395 nm) are shown. The starting points of each larval trajectory are set to 0. **(D)** Vertical swimming in wild-type (WT) and mutant (NOSΔ11 and NOSΔ23) larvae at three-day-old stimulated with UV (395 nm) light from side, blue (488 nm) light from top and UV (395 nm) light from top. The data are shown in 30 s bins. **(E)** Vertical swimming in larvae treated with NOS inhibitors, L-NAME at three-day-old stimulated with UV (395 nm) light from side, blue (488 nm) light from top and UV (395 nm) light from top. The data are shown in 30 s bins.](../figures/Fig3.png){width="100%"}

**Two unconventional guanylyl cyclases are expressed in the cPRCs**

NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PF00211). NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production. Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit [@Veraszto2017]. Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020]. To search for NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs **(Figure 4---figure supplement 1)**. To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan species and 2 choanoflagellate species. We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup. In agreement with Moroz et al. [@Moroz2020],we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans. The 15 *Platynereis* sequences were parts of several major and ancient clades of the tree **(Figure 4---figure supplement 1)**. To characterise the expression of NIT-GCs, we using previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017]. Among the 15 NIT-GCs, two showed high and specific expression in the cPRCs and one was expressed in the INNOS cells **(Figure 4---figure supplement 1)**. In the single-cell data, we could identify the cPRCs by the specific expression of c-opsin1 and a pedal-peptide2 neuropeptide precursor (MLD proneuropeptide) [@Arendt2004; @williams2017], which have previously been described as cPRC markers **(Figure 4---figure supplement 2A)**. The INNOS cells were identified by NOS expression and their spatial mapping. We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequences, NIT-GC1 and NIT-GC2. To confirm the single-cell data, we first carried out in situ HCR with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes showed very clear and specific expression in the four cPRCs, as confirmed by co-labeling with an acetylated α-tubulin antibody and with an HCR probe against *MLD proneuropeptide* **(Figure 4A,B and Figure 4---figure supplement 2A-C)**. To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins. In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus **(Figure 4C)**. Co-immunostaining with the rabbit NIT-GC1 and the rat NOS antibodies revealed that both proteins were localised in close proximity in the neurosecretory plexus **(Figure 4---figure supplement 2D)**. In contrast, NIT-GC2 strongly and specifically labelled the large sensory ciliary region of the cPRCs **(Figure 4D)**. These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

@@

To investigate how NIT-GCs-mediated NO signalling is involved in cPRC activation, the calcium sensor GCaMP6s was ubiquitously expressed in larvae and calcium imaging in the cPRC during UV/violet stimulation. As we have shown previously, a 20-sec local stimulation of the ciliated part of the cPRC with 405 nm light lead to a transient decrease in cPRC calcium levels, likely due to the hyperpolarisation of the cPRCs [@Veraszto2018]. After 20-sec, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus - a response that may involve depolarisation. This activation phase occurs after the end of the stimulation and may be due to a slower neuroendocrine feedback [@Veraszto2018]. First, to determine whether NO mediates such a feedback, we measured calcium levels in *NOS*-mutant larvae. While we detected the inhibition phase in cPRC in *NOS*-mutants, this was not followed by activation and calcium levels remained low **(Figure 4E,F)**. These results revealed an essential requirement for NO signalling in the second phase of the cPRC response. Next, to characterise the role of NIT-GCs in changes of calcium levels in cPRC, calcium imaging was performed when each *NIT-GC* gene was knocked down using morpholino **(Figure 4---figure supplement 2E,F)**. In NIT-GC1 morphant larvae, 405 nm light stimulation caused a decrease in calcium levels in the cPRC, but the increase in calcium levels that occurred after the light stimulation was stopped was strongly suppressed **(Figure 4G)**. This result is similar to the response of NOS mutants in cPRC, suggesting that NIT-GC1 mediates depolarising responses after UV stimulation of cPRC by NO and may be involved in extracellular signalling. Interestingly, in NIT-GC2 morphant larvae, an increase in calcium levels occurred immediately after light stimulation **(Figure 4H)**. The different responses of cPRC upon knockdown of NIT-GC1 and NIT-GC2 suggest that there may be different intracellular signalling mechanisms associated with their respective localisations. We focused our analysis on NIT-GC1 because it was localised to cPRC neuroaxons within the neuroplexas region and is involved in extracellular signalling after UV stimulation. It is generally known that depolarisation in the PRC involves elevated cGMPs in the photoreceptors. We therefore used an in vitro system in which cultured cells expressed NIT-GC1 to assess the cGMP activity of NIT-GC1.

![**Figure 4. cPRC-specific expression of NIT-GC1 localises in close proximity to NOS in the ANS region.** **(A, B)** Expression analysis of the NIT-GC1 (C) and NIT-GC2 (D) gene (magenta) using in situ HCR. Co-staining image using acetylated α-tubulin antibody (ac α-tub: green). **(C, D)** Localisation analysis using NIT-GC1 and NIT-GC2 antibodies. Green shows co-staining with acetylated α-tubulin antibody (ac α-tub). **(E, F)** Changes over time in GCaMP6s intensities in cPRC when stimulated with UV in WT (dark yellow) and NOS mutant (E: NOSΔ11/Δ11, F: NOSΔ23/Δ23) (light pink) larvae. **(G, H)** Changes over time in GCaMP6s intensities in cPRC when stimulated with UV in NIT-GC1 (G: NIT1) and NIT-GC2 (H: NIT2) knockdown by morpholino (MO1 and 2) larvae. **(I)** The domain structure of NIT-GC1. Below shows the mutant NIT-GC1 (ΔNIT-GC1) with deletion of the NIT domain. Grey squares indicate the transmembrane region (TM). **(J)** A schematic of the NIT-GC1 assay; COS7 cells were used to express NIT-GC1 and the cGMP indicator Green cGull, and the NO donor SNAP and DMSO as control. Green cGull shows cGMP-dependent green fluorescence. **(K-N)** The intensity changes in cGMP fluorescence over time for the four conditions are shown. Gray traces show the optical responses to individual intensities (n \> 6). The individual intensities are normalized (ΔF/F0). The dark line shows the mean value and the band the standard error range of the mean. The respective solution was acted on 2 min after the start of imaging (grey bars).](../figures/Fig4.png){width="100%"}

@@

NIT-GC1 has a highly conserved cyclase domain and activation of NIT-GC1 is expected to produce cGMP **(Figure 4I)**. Recently, a synthetic protein called Green cGull was developed as a cGMP indicator that can be used in cultured cell systems [@Matsuda2017]. We co-expressed Green cGull and NIT-GC1 using the 2A self-cleaving peptide in cultured cells COS-7 (monkey kidney) and tested the intracellular dynamics of cGMP in the activation of NIT-GC1 **(Figure 4J)**. When the NO donor SNAP was applied to the NIT-GC1 and Green cGull co-expression cells, the intracellular cGMP levels was significantly increased compared to when NIT-GC1 was not expressed **(Figure 4K,L)**. The increase in fluorescence intensity also did not occur when DMSO was applied **(Figure 4M)**. To confirm whether this increase in intracellular cGMP levels is NIT domain-dependent response of NIT-GC1, mutant form of NIT-GC1 without the NIT domain was expressed and SNAP was applied **(Figure 4I)**. As a result, the increase in fluorescence intensity was significantly reduced **(Figure 4N)**. These results indicate that NIT-GC1 produces cGMP via the NIT domain. In bacteria, it has been proposed that the NIT domain regulates cellular functions in response to changes in extracellular and intracellular nitrate and nitrite concentrations. Our results suggest that NIT-GC1 may be sensitive to NO. This could be a possible feedback pathway in which NIT-GC1 expressed in cPRC is activated by NO produced by INNOS. In order to further explore the role of NO in the cPRC circuit, we therefore compared calcium imaging of postsynaptic cells (INNOS, INRGW, Ser-h1 and MC) in the cPRC circuit during UV stimulation of cPRC between WT and NOS mutant larvae.
